Targeted interrogation of mutational hotspots in rheumatoid arthritis

### 2024-03-05 scomatic/reports/driver_coverage.Rmd
tinkering with subset_bams in scomatic.nf

---
title: "Driver coverage"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  drivers: NULL
  ref_cds: NULL
  cache_dir: './driver_coverage_cache/'
  rerun: true
---

@@

}, file = 'cov.rds', rerun = params$rerun)
```

We also load the `BSgenome` and `TxDb` objects.

@@

```{r gencov}
plot_cov <- function(cov, gene, var, expand_features_of_interest = T) {

@@

plot_cov(cov, 'ARID2', 'celltype') 
plot_cov(cov, 'ARID2', 'celltype', F)
plot_cov(cov, 'ARID2', 'id')
plot_cov(cov, 'ARID2', 'id', F)
```

### Annotate exons and introns

@@

@@

>>>>

---
title: "Driver coverage"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  drivers: NULL
  mutations: NULL
  ref_cds: NULL
  cache_dir: './driver_coverage_cache/'
  rerun: true
---

@@

}, file = 'cov.rds', rerun = params$rerun)

# read in driver mutations from LCM-WES
muts <-
  readr::read_tsv(params$mutations)
```

# GenVisR plots

We load the `BSgenome` and `TxDb` objects.

@@

```{r gencov}
# function: use gencov
plot_cov <- function(cov, gene, var, expand_features_of_interest = T) {

@@

# plot
names(cov) %>%
  purrr::walk(~ plot_cov(cov, .x, 'celltype') %>% print())
names(cov) %>%
  purrr::walk(~ plot_cov(cov, .x, 'celltype', F) %>% print())
names(cov) %>%
  purrr::walk(~ plot_cov(cov, .x, 'id') %>% print())
names(cov) %>%
  purrr::walk(~ plot_cov(cov, .x, 'id', F) %>% print())  
```

# My plots

## Annotate exons and introns

@@

## Plot driver coverages

@@

## Mean_coverage

### 2024-03-05 scomatic/reports/targeted_mutation_calling.Rmd
tinkering with subset_bams in scomatic.nf

---
title: "Targeted interrogation of mutational hotspots in rheumatoid arthritis"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  hotspots: NULL
  ref_cds: NULL
  cache_dir: './driver_coverage_cache/'
  rerun: true
---

@@

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  dpi = 300, 
  cache.path = params$cache_dir)

@@

```{r mut_hotspots}
#params<-list(drivers = 'data/driver_genes/driver_gene_coords_for_coverage.bed',hotspots='/lustre/scratch125/casm/team268im/lh22/synovium_nanoseq/ex_t_dndsout/exome/All_dnds_annot_muts.tsv')
drivers <-
  readr::read_tsv(params$drivers, col_names = c('chr', 'start', 'end', 'gene')) %>%
  dplyr::mutate(chr = gsub('chr', '', chr))
hotspots <-
  readr::read_tsv(params$hotspots)
gr <-
  GRanges(seqnames = drivers$chr,
          ranges = IRanges(start = drivers$start,
                           end = drivers$end))
```

# BAMs from snRNA-seq data

We read in reads in the driver regions

```{r scanbam}
bam_path <-
  '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/BRI_401/celltype_bams/BRI_401.B_cell_plasma_cell.bam'
bam_file <- Rsamtools::BamFile(bam_path)
bam_params <- 
  Rsamtools::ScanBamParam(which = gr, what = Rsamtools::scanBamWhat())
bam <- Rsamtools::scanBam(bam_file, param = bam_params)
```

>>>>

---
title: "Targeted interrogation of mutational hotspots in rheumatoid arthritis"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  drivers: NULL
  hotspots: NULL
  bam_mappings: NULL
  cache_dir: './target_mutation_calling_cache/'
  rerun: true
---

@@

# chunk options
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  dpi = 300, 
  cache.path = params$cache_dir)

@@

```{r mut_hotspots}
# read in bam mappings, get celltype-split bams only
mappings <-
  readr::read_csv(params$bam_mappings) %>%
  dplyr::filter(!is.na(celltype))

# read in drivers
drivers <-
  readr::read_tsv(params$drivers, col_names = c('chr', 'start', 'end', 'gene')) %>%
  dplyr::mutate(chr = gsub('chr', '', chr))

# read in mutations in drivers
hotspots <-
  readr::read_tsv(params$hotspots) %>%
  dplyr::mutate(mut_pos = pos,
                chr = gsub('chr', '', chr)) %>%
  dplyr::group_by(chr, mut_pos, gene, ref_cod) %>%
  # summarise each position
  dplyr::summarise(n = dplyr::n(),
                   id = paste(id, collapse = ','),
                   mut_cod = paste(mut_cod, collapse = ','),
                   impact = paste(impact, collapse = ',')) %>% 
  dplyr::group_by(dplyr::across(everything())) %>%
  # extend pileup window 20bp either side of mutated position
  dplyr::reframe(pos = (mut_pos - 20):(mut_pos + 20))

# save as a bed
hotspots %>%
  dplyr::transmute(chr = paste0('chr', chr), start = pos - 1, end = pos) %>%
  readr::write_tsv(
    'data/driver_genes/lcm_wes_mutations_20bp_window.bed',
    col_names = F)

# get ref allele at each position
refs <- xfun::cache_rds({
  refs <-
    system('bedtools getfasta -fi /lustre/scratch125/casm/team268im/al28/bed_ref/GRCh38_full_analysis_set_plus_decoy_hla_genome.fa -bed data/driver_genes/lcm_wes_mutations_20bp_window.bed -tab', intern = T) %>%
    paste(collapse = '\n') %>%
    readr::read_tsv(col_names = c('pos', 'ref')) %>%
    tidyr::separate_wider_delim(cols = pos, delim = stringr::regex(':|-'), names = c('chr', 'start', 'end')) %>%
    type.convert(as.is = T) %>%
    dplyr::transmute(chr = gsub('chr', '', chr), pos = end, ref) %>%
    dplyr::distinct()
  refs
}, file = 'refs.rds', rerun = params$rerun)
```

# Pileup of BAMs from snRNA-seq data

```{r pileup}
# bam2r filtering arguments
min_phred <- 30
sam_flag <- 3844
mapq <- 25
nucleotides <- c("A", "T", "C", "G", "-", "INS", "a", "t", "c", "g", "_", "ins")

# list of id - ct - bam
bams <-
  mappings %>%
  {split(., .$id)} %>%
  purrr::map(function(df) {split(df$bam, df$celltype)})

# get bases at sites
sites <- xfun::cache_rds({

  sites <-
    bams %>% 
    purrr::map(function(id) {
      id %>% 
        purrr::map(function(ct_bam) {

          hotspots %>%
            purrr::pmap(function(chr, pos, mut_pos, gene, ...) {
              deepSNV::bam2R(
                ct_bam,
                chr = chr,
                start = pos,
                stop = pos,
                q = min_phred,
                mask = sam_flag,
                mq = mapq) %>%
                tibble::as_tibble() %>%
                dplyr::mutate(chr = chr, pos = pos, mut_pos = mut_pos, gene = gene) %>%
                dplyr::select(chr, pos, mut_pos, gene, dplyr::all_of(nucleotides))
            }) %>%
            dplyr::bind_rows() %>%
            dplyr::rowwise() %>%
            dplyr::mutate(
              total = sum(dplyr::across(dplyr::all_of(nucleotides)))) %>%
            dplyr::relocate(dplyr::all_of(nucleotides), .after = dplyr::last_col())

        }) %>% dplyr::bind_rows(.id = 'celltype')
    }) %>% dplyr::bind_rows(.id = 'id')

  # return
  sites

}, file = 'sites.rds', rerun = params$rerun)

# add refs, compare
muts <-
  sites %>%
  dplyr::left_join(refs) %>%
  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                      values_to = 'nt_n_reads', names_to = 'nt') %>%
  # # ignore indels for now
  # dplyr::filter(!nt %in% c('-', '_', 'INS', 'ins')) %>%
  # get number of ref and non-ref alleles at each position
  dplyr::mutate(n_ref = ifelse(ref == toupper(nt), nt_n_reads, 0),
                n_non_ref = ifelse(ref != toupper(nt), nt_n_reads, 0)) %>%
  dplyr::group_by(id, celltype, chr, pos, mut_pos, gene, total, ref) %>%
  dplyr::summarise(n_ref = sum(n_ref), n_non_ref = sum(n_non_ref)) %>%
  dplyr::mutate(vaf = n_non_ref / total)

# plot positions
unique(sites$gene) %>%
  purrr::walk(function(i) {
    p <-
      sites %>%
      dplyr::filter(gene == i) %>%
      dplyr::left_join(refs, by = c('chr', 'pos')) %>%
      tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                          values_to = 'nt_n_reads', names_to = 'nt') %>%
      dplyr::mutate(
        strandless_nt = 
          ifelse(nt %in% c('-', '_'), 'DEL', toupper(nt)),
        is_ref = ifelse(strandless_nt == ref, 'ref', 'alt'),
        strandless_nt = paste0(strandless_nt, '_', is_ref) %>% 
          forcats::fct_relevel(unique(c(paste0(topupper(nucleotides), '_ref'),
                                        paste0(toupper(nucleotides), '_alt'))))) %>%
      ggplot(aes(x = pos, y = nt_n_reads, fill = strandless_nt)) + 
      geom_vline(aes(xintercept = mut_pos)) +
      geom_col(position = 'stack') +
      theme_bw() +
      theme(axis.text.x = element_blank()) +
      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free_x') +
      scale_fill_manual(values = c(A_ref = 'darkgreen', A_alt = 'lightgreen',
                                   C_ref = 'darkblue', C_alt = 'lightblue',
                                   G_ref = 'darkorange3', G_alt = 'peachpuff',
                                   T_ref = 'darkred', T_alt = 'salmon',
                                   INS_alt = 'black', DEL_alt = 'grey')) +
      ggtitle(i)
    print(p)
    })
```
